---
title: "Flight_Delay_Prediction_using_R"
author: "Joseph Vinay Kumar POLANKI"
date: "`r Sys.Date()`"
output: word_document
---

## do it with both NA values in the DATA and removing NA values in DATA

# Setting up the knitr file

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

## Abstract

## Research Questions

## DATA

##Loading DATA into rmd file from drive.

```{r Loading_Data, echo= TRUE}
#Loading and Inspecting the DATA
install.packages("readr")
library(readr)
Flight_DATA <- read.csv("G:/My Drive/CIS 663 R programming/ATC_R_Project/DATA/air_traffic_main_Data.csv")
```

## Viewing first and last few rows.

```{r  Viewing_Data, echo = TRUE}
head(Flight_DATA)
tail(Flight_DATA)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
install.packages("ggplot2")
library(ggplot2)
```

```{r }
install.packages("dplyr")
library(dplyr)
glimpse(flight_data)

```

```{r  echo=TRUE }
V <- is.na(Flight_DATA)   #V <- sum(is.na(Flight_DATA))

y <- sum(V)

y

#Data set have 120 rows of NA values combined in all columns.

```

```{r echo = TRUE,warning=FALSE}

install.packages("tidyr")
library(tidyr)

# creating a copy original FLIGHT_DATA

Copy_DATA <- Flight_DATA

#omitting all the NA Values to see no of rows and Columns

Copy_Flight_DATA <- Copy_DATA %>% drop_na()


# checking wether they are IDENTICAL

identical(Copy_Flight_DATA,Flight_DATA)

# understanding the change
dim(Flight_DATA)
dim(Copy_Flight_DATA)

```

```{r }

# Combine the two columns into one
flight_data <- Flight_DATA %>%
 mutate(combined_delay = paste(security_delay, weather_delay, carrier_delay, nas_delay, late_aircraft_delay, sep = "_"))

```
## if combined delay is not converted to numeric throwing an error.

```{r }
combined_delay <- as.numeric(flight_data$combined_delay)

```

```{r }

# Plotting histogram of combined delay minutes
ggplot(flight_data, aes(x = weather_delay)) +
  geom_histogram(fill = "skyblue", color = "white") +
  labs(title = "Distribution of Flight Delay Minutes",
       x = "Combined Delay Minutes",
       y = "Frequency")


```

```{r }
delay_by_airport <- Flight_DATA %>%
group_by(airport) %>%
summarize(AVG_DELAY = mean(, na.rm = 'TRUE'))

```

# Plotting delay by airport

```{r }
ggplot(delay_by_airport, aes(x = airport, y = avg_delay)) + geom_bar(stat = "identity", fill = "skyblue", color = "black") + labs(title = "Average Delay by Airport", x = "Airport", y = "Average Delay (minutes)")
```

```{r }

delay_by_airport <- Flight_DATA %>%
dplyr:::summarize.grouped_df(., AVG_DELAY = mean(delay_minutes, na.rm = "TRUE"))
summarize(AVG_DELAY = mean(delay_minutes, na.rm = TRUE))

```





```{r    echo = FALSE}
library(ggplot2)

ggplot(Flight_DATA, aes(x = arr_delay)) +
  geom_histogram(binwidth = 5000, fill = "coral", color = "orange") +
  labs(title = "Distribution of Flight Delays", x = "Arrival Delay (minutes)", y = "Frequency")
```



```






















