---
title: "Flight_Delay_Prediction_Using_R"
author: "Joseph Vinay Kumar POLANKI"
date: "`r Sys.Date()`"
output: word_document
---

## Youtube Links

<https://www.youtube.com/watch?v=tp_BG5wDeVU> (hist)

<https://www.youtube.com/watch?v=BTDSzqHW_5E> <https://www.youtube.com/watch?v=wsi0jg_gH28&t=4s> (linear)

<https://www.youtube.com/watch?v=dWjFReizadM&t=1s> (reg models viz)

<https://www.youtube.com/watch?v=1MHX1s5bb6w> (shiny)

<https://www.youtube.com/watch?v=SGOmGAwsxME> <https://www.youtube.com/watch?v=HbkgAUOYvCY> (flex) <https://www.youtube.com/watch?v=O3CgrEwTg1k> (flex)

<https://www.youtube.com/watch?v=q8eR2suCyGk&t=2s> (NA)

<https://www.youtube.com/watch?v=O42iklO8uYU> (NA)

## do it with both NA values in the DATA and removing NA values in DATA

when tried it is giving an error or NA as mean of the data for a column.

## to see the colours of ggplot

## Packages to use

(caret) (randomForest) (rpart) (e1071) ggplot dplyr readr stargazer

# Setting up the knitr file

```{r setup, include= TRUE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

## Abstract

## DATA

DATA for the project was downloaded from kaggle, which includes historical flight data, such as carrier information, Flight Schedules and delay statistics sourced from reputable aviation Databases.

##Loading DATA into rmd file from drive. Rows: 3351 Columns: 21

```{r Loading_Data, echo= TRUE}
#Loading and Inspecting the DATA
install.packages("readr")
library(readr)
Flight_DATA <- read.csv("G:/My Drive/CIS 663 R programming/ATC_R_Project/DATA/air_traffic_main_Data.csv")
```

## Viewing first and last few rows.

```{r  Viewing_Data, echo = TRUE}
head(Flight_DATA)
tail(Flight_DATA)
```

## Installing ggplot and dplyr packages for Data Exploration.

```{r, packages , echo=TRUE}
install.packages("ggplot2")
library(ggplot2)


install.packages("dplyr")
library(dplyr)
```

```{r }

# Install and load the color space package to see the colors available in "ggplot"
install.packages("colorspace")
library(colorspace)

# Display the available colors.
windows()
colors()

```

```{r , Understanding the DATA,  echo= FALSE}

#Basic Information and Structure of the Data

glimpse(Flight_DATA)

#Summary statistics of the data

summary(Flight_DATA)

```

```{r  echo=TRUE }

V <- is.na(Flight_DATA)   #V <- sum(is.na(Flight_DATA))
y <- sum(V)
y

#Data set have 120 rows of NA values combined in all columns, 8 rows in each column.

```

## Install tidyr package and drop rows with NA values.

```{r,  echo = FALSE}

install.packages("tidyr")
library(tidyr)

# creating a copy original FLIGHT_DATA

Copy_DATA <- Flight_DATA

#omitting all the NA Values to see no of rows and Columns

Copy_Flight_DATA <- Copy_DATA %>% drop_na()


# checking whether they are IDENTICAL

identical(Copy_Flight_DATA,Flight_DATA)

# understanding the change
dim(Flight_DATA)
dim(Copy_Flight_DATA)

```

## Exploratory DATA Analysis (EDA)

```{r    echo = FALSE}
library(ggplot2)

# Displaying the arrival Flight delays. 

ggplot(Flight_DATA, aes(x = arr_delay)) +
  geom_histogram(binwidth = 15, fill = "coral", color = "orange") +
  labs(title = "Distribution of Flight Delays", x = "Arrival Delay (minutes)", y = "Frequency Of Flights")
```

Most of the flights coming on time with minimal or no delays but few of them are late arrivals

```{r    echo = FALSE}
library(ggplot2)

#Delays by Carrier

ggplot(Flight_DATA, aes(x = carrier_delay)) +
  geom_histogram(binwidth = 15, fill = "steelblue", color = "green") +
  labs(title = "Distribution of Flight Delays", x = "Arrival Delay (minutes)", y = "Frequency Of Flights")
```

```{r }

ggplot(Flight_DATA, aes(x = weather_delay)) +
  geom_histogram(binwidth = 15, fill = "darkgray", color = "brown") +
  labs(title = "Distribution of Flight Delay Minutes",
       x = "weather Delay Minutes",
       y = "Frequency Of Flights")


```

```{r }


Flight_DATA %>%

group_by(airport) %>%

summarise(mean_delay = mean(arr_delay, na.rm =TRUE)) %>%

ggplot(aes(x = reorder(airport,mean_delay), y= mean_delay)) +
geom_bar (stat = "identity", fill = "purple") +
  coord_flip() +
  labs (title = "AVG DELAY BY AIRPORT", X = "Frequency Of Flights", y = "AVG Delay In Minutes")


```

```{r }

Flight_DATA %>%
  group_by(airport) %>%
  summarise(mean_delay = mean(arr_delay, na.rm = TRUE)) %>%
  top_n(25, mean_delay) %>%  # Select top 25 airports by mean delay
  ggplot(aes(binwidth = 15, x = reorder(airport, mean_delay), y = mean_delay)) +
  geom_bar(stat = "identity", fill = "purple") +
  coord_flip() +
  labs(title = "Top 25 Airports by AVG Delay", x = "Airport", y = "AVG Delay In Minutes")
```










## END FOR TODAYS.





```{r }
Flight_DATA %>%
  group_by(airport) %>%
summarise(mean_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(mean_delay)) %>%
top_n(10)


```

  

  
  
  

```{r }
# Combine the two columns into one
flight_data <- Flight_DATA %>%
 mutate(combined_delay = paste(security_delay, weather_delay, carrier_delay, nas_delay, late_aircraft_delay, sep = "_"))

```

## if combined delay is not converted to numeric throwing an error.

```{r }
combined_delay <- as.numeric(flight_data$combined_delay)

```

```{r  }
delay_by_airport <- Flight_DATA %>%
group_by(airport) %>%
summarise(AVG_DELAY = mean(, na.rm = 'TRUE'))

```

# Plotting delay by airport

```{r }
ggplot(delay_by_airport, aes(x = airport, y = avg_delay)) + geom_bar(stat = "identity", fill = "skyblue", color = "black") + labs(title = "Average Delay by Airport", x = "Airport", y = "Average Delay (minutes)")
```

```{r }

delay_by_airport <- Flight_DATA %>%
dplyr:::summarize.grouped_df(., AVG_DELAY = mean(delay_minutes, na.rm = "TRUE"))
summarize(AVG_DELAY = mean(delay_minutes, na.rm = TRUE))

```
