---
title: "Flight_Delay_Prediction_using_R"
author: "Joseph Vinay Kumar POLANKI"
date: "`r Sys.Date()`"
output: word_document
---



## Youtube links


https://www.youtube.com/watch?v=v048ru5v0BE (R_markdown file editing)

https://www.youtube.com/watch?v=0KfoJ3NiFK4 (Intro to STAT tests)

https://www.youtube.com/watch?v=lyXd3zCCxto (gganimate for presentation)

## Regression Analysis

https://www.youtube.com/watch?v=66z_MRwtFJM(Linear)

https://www.youtube.com/watch?v=mno47Jn4gaU (Multiple Regression(STATQUeST))

https://www.youtube.com/watch?v=wsi0jg_gH28 (Linear (STATQUEST))

https://www.youtube.com/watch?v=q8eR2suCyGk (NA VALUES)


https://www.youtube.com/watch?v=dWjFReizadM (Regression Models)

https://www.youtube.com/watch?v=AR6sLpcVcSU (prediction using regression)

https://www.youtube.com/watch?v=O42iklO8uYU  (Missing Values in R)

https://www.youtube.com/watch?v=UdndWOsRm5E  (subsetting missing values in R)

https://www.youtube.com/watch?v=uaCK6wFcX3I   (subsetting missing values in R)

https://www.youtube.com/watch?v=1MHX1s5bb6w (shiny for Dashboard)

https://www.youtube.com/watch?v=fkqD9kcvCkU  (dashboard creation in R)

https://www.youtube.com/watch?v=RGXUkXNrVxU  (Flexdashboards)

https://www.youtube.com/watch?v=jxsKUxkiaLI (shiny web apps)










# Setting up the knitr file
Rendered Document after knitting of the "rmd" file shows the code and output
where as the warnings generated during the code execution is excluded from the code
below but the code in this particular code chunk is not displayed as this code chunk is used to set global options.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

## Abstract



## Research Questions


## DATA
setting working directory to avoid errors while saving binary files

```{r echo=TRUE}
setwd("G:/My Drive/CIS 663 R programming/ATC_R_Project/DATA")
```


##Loading DATA
Loading Data from drive and installing "readr" package.

```{r Loading_Data, echo=TRUE}
#Loading and Inspecting the DATA
install.packages("readr")
library(readr)
Flight_DATA <- read.csv("G:/My Drive/CIS 663 R programming/ATC_R_Project/DATA/air_traffic_main_Data.csv", header = TRUE, stringsAsFactors = FALSE)

```

## Viewing first and last few rows.

```{r  Viewing_Data, echo = FALSE}
head(Flight_DATA)
tail(Flight_DATA)
```


```{r summaries, echo = TRUE }

is.na(Flight_DATA)

summary(Flight_DATA)

grouping(Flight_DATA$month)

```


```{r, echo=TRUE}

v <- c(Flight_DATA$arr_flights)




```




```{r}

```







